---
title: "A very quick introduction to vegetation assessments"
subtitle: 'Conservation & Land Mangement Training 2024'
author: "Reed Benkendorf, Chris Woolridge, Jacob Zeldin & Matt Evans"
date: "June 5^th^ 2024"
output:
  ioslides_presentation:
    widescreen: true
    logo: ../images/cbg_logo.jpg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(dpi = 300) 
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
knitr::opts_chunk$set(fig.align="left") 
```

```{r load libraries}
library(tidyverse)
library(rstatix)
library(iNEXT)
library(vegan)
```


## What is Floristic Quality? 
<font size="4">

> _'... tolerance of anthropogenic disturbance and exclusivity to remnant habitats are the only validated criteria for deﬁning FQA.'_
> 

> _'...FQA conveys two things about high conservative species: (1) All else being equal, they have greater conservation value, and (2) they reﬂect a site’s history of minimal disturbance and degradation.'_
>
> `r tufte::quote_footer('--- Spyreas 2019')`

</font>

- Useful for identifying high-quality remnant habitat  
- Essentially a format for prioritizing land management decisions, especially active management 
- We know we cannot restore everywhere, preventing degradation is key, we can shift certain activities to areas of less conservation values  

## What do C-values mean? 

- Vary by each administrative unit they are generated for  
- Reflect the *purpose* of monitoring  

<font size="4">

> _'... the coefficient applied to each taxon represents a rank based on observed
behavior and patterns of occurrence in Illinois plant communities and our confidence that a taxon is remnant
(natural area) dependent._   

> _Species given a C value of **0-1** are taxa adapted to severe disturbances, particularly
anthropogenic disturbances..._   

> _Species with a C value of **2-3** are associated with somewhat more stable, though degraded,
environments._   

> _Species with coefficients **4-6** include many dominant or matrix species for several habitats;
they have a high consistency of occurrence within given community types._ 

> _Species with C values **7-8** are taxa we associate mostly with natural areas, but that can be found persisting where the habitat has been degraded somewhat._    

> _Species with coefficients **9-10** are considered to be restricted to high-quality natural areas.'_    
> 
> \hfill `r tufte::quote_footer('--- Taft et al. 1997')`

</font> 

## What do we mean by Species Richness? 


https://www.webpages.uidaho.edu/veg_measure/index.htm great resource on vegetation ecology in natural resource management!   

## Plot Layout 

- Species Richness  
  - Rectangular plots capture more species (reviewed in Elzinga et al. 2001)  
    - Cross more micro-habitats  
    - More likely to 
- Species Abundance  
  - Foliar cover / Rooting Cover  
    - How much 
  - Indicator species, frequency of certain grass species; nested quadrats
    - strength of frequency: unlikely to vary via climate or grazing  

## Reading Plots  
- Boundary Decisions
  - Are plants rooted? 
  - Which side of the plots are plant rooted in?  
- Order to reading?   
  - Nested quadrats go from smallest to largest plot
    - Only record *new* species along the way 
- Avoiding double counting
  - search methodically following a single direction  

## Management Decisions - Species Richness

- Ecosystem resilience often found to relate to the number of species in an area  
  - Loss of species may allow proliferation of invasive species  

## Management Decisions - FQA  

- Should areas with more remnant species be prioritized for active management? *and why* 
  - Some species cannot be 'added' back in via restoration  
  - These areas maybe managed to reduce further anthropogenic impacts, e.g. invasive removal, disallow certain recreation activities. 
  - Some areas with higher C-values likely to still harbor rare-species, impetus to survey. 

- Use Mean C, or FQI to detect these areas.  

## Management Decisions - Species Cover 

- Cover of species naturally ebb and flow  
- Certain management actions can increase cover of some species, while vastly decreasing cover of others  
- Many of the decreasing species may be important for wildlife, or ecosystem function (e.g. water retention, soil stability)  

## Management Decisions - Vegetation Similarity 

- If we want to *restore* an area, what target area should we use to guide our decisions?  
- Can we expect areas with similar vegetation to respond similarly to management actions? 
  - e.g. fire, grazing cattle/sheep/goats, removal of invasives? 

## Do people really do this?  

- Yes, see Benkendorf & Holsinger 2023 for a real life example  
  - They use plot based species richness data to determine which factors most affect status of natural areas
  - ca. 860,000 acre area, use spatial statistics and regression technicques to identify features correlating with high FQI  
  - Decision: 
    - Do we restore areas at less than 7,000 feet? 
    - Or do we climate proof ourselves and focus higher?  
  - Surprising finding
    - Areas near human habitation doing pretty well...  


## Analyses 

```{r}

```

## Analysis 1 - Are we capturing a representative amount of the species diversity? 

- Not all species are equally common, and detection of locally rare species is difficult  
- We can assess sampling completion using rarefaction
- If the number of species (Y axis), increase greatly via extrapolation, then we should sample more plots  
- Helps us determine whether more rare species, likely with higher C values, were missed by sampling  
- No hypothesis 

```{r Estimating Species Richness of the two areas}

data(ciliates)

rarified <- iNEXT(ciliates, q=0, datatype = "incidence_raw")
ggiNEXT(rarified, type=1, facet.var = "Assemblage") +
  labs(
    title = 'Rarefaction curve of Species Richness', 
    x = 'Plots', y = 'Species Richness (q = 0)') + 
  theme_minimal() + 
  theme(legend.position = 'bottom') 
```


## Analysis 2 - Are the two sites notably different? 

- FQI reflects the history of disturbance of a site, the sites disturbance histories vary  
- Hypothesis: the *Swamp white Oak section* has less historic anthropogenic disturbance  
- Null Hypothesis: the FQI of the two sites do not differ  
- Two-sided Wilcoxon Signed Rank Test, data are unpaired   

```{r Do the two areas differ?}
fqi_p <- wilcox_test(plotData, FQI ~ Site, paired = FALSE)
fqi_r <- wilcox_effsize(plotData, FQI ~ Site, paired = FALSE)

coc_p <- wilcox_test(plotData, FQI ~ Site, paired = FALSE)
coc_r <- wilcox_effsize(plotData, FQI ~ Site, paired = FALSE)

## Generate Box Plots of data and results  

tg <- ToothGrowth %>% 
  filter(dose < 2) %>% 
  mutate(dose = as.factor(dose))

fqi_p <- wilcox_test(tg, len ~ dose, paired = FALSE)
fqi_r <- wilcox_effsize(tg, len ~ dose, paired = FALSE)

fqi_r$effsize
round(fqi_p$p, 5)

fqi <- ggplot(tg, aes(x = dose, y = len, fill = dose)) + 
  geom_boxplot(notch = TRUE) + 
  theme_minimal() + 
  labs(
    title = 'Comparision of FQI values between Sites', 
    y = 'Floristic Quality Index', x = 'Site', 
    caption = bquote(
      paste(
        'Strong evidence (', 
        italic(p), ' = ', 
        .(round(fqi_p$p, 5)),
        ') was found that the two sites (', 
        italic(n), '= ', 
        .(fqi_p$n1), 
        ') differ greatly (effect size = ',
        .(round(fqi_r$effsize, 2)),
        ') in their FQI scores.')
      )
    ) + 
   theme(
     plot.title = element_text(hjust = 0.5),
     plot.caption = element_text(hjust = 0) 
     )

```

## Analysis 3 - What Quadrat size can we use to determine frequency of Species of interest this year? 

- It can be difficult to determine the appropriate plot size for monitoring  
- Nests allow us to try different options readily  
- For expedited monitoring carried out across large domains, finding an optimal size is important  
- We want a species to show up in 20-80% of plots of various sizes  

## Analysis 4 - Are there microhabits more similar across than between sites? 

- Some plots may have more similarities to plots at the other site  
- This may indicate each site has pockets in similar states  
- No hypothesis 


```{r transform nested data into quasi abundance data}

# nmds require abundance data, these are best collected via occular estimates 
# of plant cover, or an objective method e.g. line or line-point intercept 

quarter <- 0.25*0.25
half <- (0.5*0.5) - quarter
mid <- (1*1) - half -quarter

vals <- data.frame(
  quarter = 1/quarter, # species found here are 16 times more 'abundant' than those in largest quadrats
  half = 1/half, # species found here are 5.3 times more 'abundant' than those in the largest quadrat.
  mid = 1/mid, # species found here are 1.3 times more 'abundant' than those in the largest quadrat.
  full = 1 # constant, they remain 1
) %>%  
  t() %>% 
  data.frame() %>% 
  rownames_to_column() %>% 
  setNames(., c('quadrat', 'constant'))


# for each quad level, divide by the number of species within it?? /
quadrat = c('quarter', 'quarter', 'half', 'half', 'half', 'mid', 'mid', 'mid', 'full', 'full', 'full', 'full', 'full')
df <- data.frame(
  quadrat = quadrat, 
  species = c(LETTERS[1:length(quadrat)])
)

df <- left_join(df, vals)
df$adjusted <- jitter( 
  df$constant * 100 / 
    sum(df$constant) # now make values add up to ~100 % for quasi cover comparision across plots. 
  ) # we can't have ties, so jitter. 


```


```{r Are the plots more similar to each other}
library(vegan)
set.seed(12)
community_matrix <- matrix(
   sample(1:100,300,replace=T), nrow=10,
   dimnames=list(paste("comm", 1:10, sep=""), paste("sp",1:30,sep="")))

example_NMDS <- metaMDS(community_matrix, k = 3, trymax = 500) # The number of reduced dimensions

treat=c(rep("Upland",5), rep("Swamp",5))

par( mfrow= c(3,1), mai = c(0.25, 0.5, 0.1, 0.1), pty = 's')
# the plots can be busy... let's build them up slowly... 
ordiplot(example_NMDS, type = "n", xaxt = 'n', annX=FALSE) # showcase the species in space   
orditorp(example_NMDS, display = "species", col="#00a6ed", air = 0.01, cex = 0.7)

ordiplot(example_NMDS, type = "n", xaxt = 'n', annX =FALSE) # showcase the two main clusters we found     
orditorp(example_NMDS, display = "species", col="#00a6ed", air = 0.01, cex = 0.7)
ordihull(example_NMDS, groups = treat, draw = "polygon", col = c('#fe7f2d', '#264653'), label=T)

ordiplot(example_NMDS, type = "n") # showcase the plots we found and relations between them 
ordihull(example_NMDS, groups = treat, draw = "polygon", col = c('#fe7f2d', '#264653'), label=F)
orditorp(example_NMDS, display = "sites", col = c(rep("#fe7f2d",5), rep("#264653",5)), air = 0.5, cex=1.1)

```

## Analysis 5 - Change in species composition across sites, and total species 

- beta diversity
- gamma diversity 